In-silico molecular enrichment and clearance of the human intracranial space

### 2024-12-19 paper/supplementary_main.tex
Finalize dispersion clean-up in supplementary.

\mer{MER: Stopped here.}

\subsection{Modelling dispersion in the CSF spaces}
\label{sec:app:dispersion}

\mer{MER: Moved the below from Results.}
\mar{To examine cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm LV}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model.}

% respiratory pulsatility
\mar{Employing the same methodology, we estimate the respiratory contribution to dispersive mixing by computing the CSF flow field driven by an inflow of $0.121\,$ml/s \cite{liu2024using} and $1\,$ml/s \cite{gutierrez2022effect} across the lateral ventricular wall and the arachnoid mater, respectively.}

CSF pulsates in the SAS and ventricular system with each cardiac and respiratory cycle, leading to an increase in effective solute diffusivity via a process known as Taylor dispersion \cite{taylor1953dispersion, watson1983diffusion}. Previous estimates of the magnitude of this effect in the CSF-filled space range from an enhancement factor of 5.8 \cite{keith2019dispersion} up to $10^4$ \cite{ray2021quantitative}. The large discrepancies in reported estimates can (at least partly) be attributed to different assumptions on representative values for local width, pressure differences and fluid velocities in the SAS, which reflect the diverse flow characteristics in the different regions of the intracranial CSF-filled space.

Taking the large spatial differences of pulsatile CSF flow into account, we estimate a spatially varying diffusion enhancement factor $R$ in the CSF-filled space $\Omega_{\rm CSF}$ by combining a computational model of cardiac- and respiration-induced pulsatile CSF flow with theoretical estimates for shear-augmented (Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion}. These estimates are are derived under simplifying assumptions on the problem geometry, the flow field and concentration profiles \cite{watson1983diffusion, keith2019dispersion}. Therefore such an estimate is heuristic; however, our computational results qualitatively agree with observed data.

Similar to the case of CSF flow driven by production introduced in \Cref{sec:csf_flow}, we employ a Stokes flow model to predict CSF velocities corresponding to the peak cardiac and respiratory blood inflow rates. In particular, we enforce a uniformly distributed inflow of $u^{\rm in}_{\rm AM}$ and $u^{\rm in}_{\rm LV}$ across the arachnoid mater ($\Gamma_{\rm AM}$) and lateral ventricular surfaces ($\Gamma_{\rm LV}$) respectively, mimicking the effect of reduced CSF volume due to blood inflow. We set $u^{\rm in}_{\rm AM} = 6\,$ml/s \cite{causemann2022human,baledent2014imaging} and $u^{\rm in}_{\rm LV}=0.31\,$ml/s \cite{vinje2019respiratory} in the cardiac case, and $u^{\rm{in}}_{\rm{AM}} = 1\,$ml/s \cite{gutierrez2022effect} and $u^{\rm in}_{\rm{LV}}=0.121\,$ml/s \cite{liu2024using} in the respiratory case, respectively.
We prescribe a no-slip boundary condition on the pial surface $\Gamma_{\rm Pia}$ and zero-traction on the spinal boundary $\Gamma_{\rm SSAS}$, making the spinal compartment the only outflow route of CSF. Thus, we compute the pulsatile flow fields by augmenting the Stokes flow model (\eqref{eq:momentum_equation} and \ref{eq:divergence_equation}) with the following boundary conditions:
\begin{equation}\label{eq:cardiac_flow_bcs}
    \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
    \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
    \bm u = \frac{u^{\rm in}_{\rm LV} \cdot \bm n}{|\Gamma_{\rm LV}|} \text{ on } \Gamma_{\rm LV}; \quad
    \bm u = \frac{u^{\rm in}_{\rm AM} \cdot \bm n}{|\Gamma_{\rm skull}|} \text{ on } \Gamma_{\rm AM}.
\end{equation}
The obtained pressure fields $p_{\rm cardiac}$ and $p_{\rm resp}$ only account for viscous forces, albeit cardiac and respiratory induced CSF flow is known to be unsteady and substantially impacted by transient inertial forces. Assuming an oscillatory flow with either the heartbeat frequency $\omega_{\rm cardiac} = 2 \pi$ or respiratory frequency $\omega_{\rm resp} = 0.5 \pi$, a mean gap width of the CSF-filled spaces of $2h=3$\,mm and CSF density $\rho=10^3\,\text{kg/m}^3$, the squared Womersley number $\alpha_i^2 = \frac{h^2 \omega_{i} \rho}{\mu}$ becomes $\alpha_{\rm cardiac}^2 = 20.2$ and $\alpha_{\rm resp}^2=5.05$, which is unsteady. Based on theoretical considerations on the ratio of oscillatory flow to steady flow impedances in a tube, we account for inertial forces in the pulsatile flow by upscaling the computed viscous pressure field by a factor of $1 + \alpha^2 / 8$ (see \cite{van1998cardiovascular}, chapter. 4.3). Thus, the upscaled pressure fields become $p^{\rm up}_{i}= (1 + \alpha_{i}^2 / 8)p_i \text{ for } i = \{\mathrm{cardiac}, \mathrm{resp}\}$.

Further assuming unsteady dispersion, we follow Sharp et al \cite{keith2019dispersion} and estimate each enhancement factor $R_{\rm cardiac}$ and $R_{\rm resp}$ from the non-dimensionalized pressure gradient 
$$R_{i} \approx dP_{i}^2 / \alpha_{i}^3, \quad  \text{where} \quad  dP=\frac{|\nabla p^{\rm up}_{i}|}{\omega_{i} \mu / h} \quad \text{for} \quad i = \{\mathrm{cardiac}, \mathrm{resp}\}.$$
Finally, we account for the non-local nature of the dispersion mechanism and smooth the resulting field by solving a Helmholtz problem with $dP+i^2/\alpha_i^3$ as the right-hand side
\begin{equation}
- 10^{-4} \Delta R_i + R_i = dP_i^2/\alpha_i^3 \quad \text{ on } \Omega_{\rm CSF} \qquad \text{ and } \quad \nabla R_i \cdot \bm n = 0, \quad \text{ on } \partial \Omega_{\rm CSF}
\end{equation}
to obtain the locally averaged, spatially varying diffusion enhancement factor $R_i$. The diffusion enhancement factor $R$ in $D^{\mathrm{eff}} = D^{\mathrm{Gad}}( 1 +R)$ is then estimated by $R \approx R_{\rm{cardiac}} + R_{\rm{resp}}$.

>>>>

\subsection{Estimating dispersion factors from pulsatile CSF flow}
\label{sec:app:dispersion}

The cardiac ($\sim$1Hz) and respiratory ($\sim$0.25Hz) cycles induce
pulsatile flow of CSF in the ventricular system and in the cranial and
spinal SAS. Pulsatile flow leads to dispersion which in turn may
enhance molecular transport~\cite{taylor1953dispersion,
  watson1983diffusion, asgari2016glymphatic, keith2019dispersion,
  ray2021quantitative, troyetsky2021dispersion}. To account for the
dispersive effects over a longer time scale (hours to days), and in
the absence of measurements or estimates of dispersion coefficients in
human CSF spaces, we adapt existing theoretical
estimates~\cite{watson1983diffusion, keith2019dispersion}. More
specifically, we compute spatially-varying dispersion enhancement
fields $R_c$ and $R_r$ (\Cref{fig:csf}G, I), associated with the
cardiac and respiratory cycles respectively, via the algorithm
presented below. These fields then contribute to the diffusion
coefficient in $\Omega_{\rm CSF}$ in~\eqref{eq:3d_pde} as $D = (1 +
R_c + R_r) D^{\rm Gad}$.
\begin{enumerate}[label=\roman*)]
\item
  To account for viscous forces, we compute spatially-varying CSF
  pressure fields $p^0_c, p^0_r$ in $\Omega_{\rm CSF}$ corresponding
  to the Stokes flow induced by the peak volumetric reduction of the
  CSF space in the respective cycle (\Cref{fig:csf}F, H). More
  precisely, we numerically solve the incompressible Stokes
  equations~\eqref{eq:stokes} equipped with the following boundary
  conditions mimicking a dilation of the brain parenchyma with the
  spinal SAS as the only route for CSF efflux:
  \begin{equation}\label{eq:pulsatility:bcs}
    \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
    \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
    \bm u = \frac{u^{\rm in}_{\rm LV} \cdot \bm n}{|\Gamma_{\rm LV}|} \text{ on } \Gamma_{\rm LV}; \quad
    \bm u = \frac{u^{\rm in}_{\rm AM} \cdot \bm n}{|\Gamma_{\rm skull}|} \text{ on } \Gamma_{\rm AM}.
\end{equation}
  In the cardiac cycle case, we set $u^{\rm in}_{\rm AM} = 6\,$ml/s \cite{causemann2022human,baledent2014imaging} and $u^{\rm in}_{\rm LV}=0.31\,$ml/s \cite{vinje2019respiratory} to solve for $p^0_c$. In the respiratory cycle case, we set $u^{\rm{in}}_{\rm{AM}} = 1\,$ml/s \cite{gutierrez2022effect} and $u^{\rm in}_{\rm{LV}}=0.121\,$ml/s \cite{liu2024using} to solve for $p^0_r$.
\item
  We also estimate the Womersley numbers $\alpha_c, \alpha_r$
  associated with the cardiac and respiratory flow patterns,
  respectively, by the definition
  \begin{equation}
    \alpha^2_i = \frac{h^2 \omega_i \rho}{\mu} \qquad i \in \{c, r\},
  \end{equation}
  with CSF density $\rho = 10^3$ kg/m$^3$, a mean CSF space width $h =
  1.5$mm, and CSF viscosity $\mu$ given in~\Cref{tab:parameters}.  For
  the cardiac cycle, we consider an angular frequency $\omega_c = 2
  \pi$, while for the respiratory cycle, we set $\omega_r = 0.5
  \pi$. The resulting (square) Womersley numbers are $\alpha_c^2 = 20.2$
  for the cardiac cycle and $\alpha_r^2 = 5.05$ for the respiratory
  cycle.
\item
  To account for inertial forces in addition to the viscous forces, we
  use the Womersley numbers $\alpha_c, \alpha_r$ to calculate upscaled
  pressure fields $p^1_c, p^1_r$ from $p^0_c, p^0_r$ as:
  \begin{equation}
    p^1_i(x) = \left ( 1 + \frac{\alpha_i^2}{8} \right ) p_i^0(x)
    \qquad \text{for } x \in \Omega_{\rm CSF}, \qquad i \in \{c, r\}.
  \end{equation}
  Note that this scaling is based on theoretical considerations on the
  ratio of oscillatory flow to steady flow impedances in a
  tube~\cite[Chap.~4.3.]{van1998cardiovascular}.
\item
  Further, assuming unsteady dispersion, we follow Sharp et
  al.~\cite{keith2019dispersion} to estimate local enhancement factors
  $S_c, S_r$ from the non-dimensionalized pressure gradients:
  \begin{equation}
    S_i(x) = \frac{1}{\alpha_i^3} \frac{1}{\omega_i \mu/h} | \nabla p_i^1(x) |
    \qquad i \in \{c, r\}.
  \end{equation}
\item
  Finally, we define the cardiac and respiratory dispersion
  enhancement factors $R_c$ and $R_r$ by smoothing $S_c$ and $S_r$,
  respectively, to account for the non-local nature of
  dispersion. Specifically, for $i \in \{c, r\}$, we define $R_i :
  \Omega_{\rm CSF} \rightarrow \R^{+}$ by solving a heuristic weighted
  Helmholtz problem over $\Omega_{\rm CSF}$ with $S_i$ as the
  right-hand side:
  \begin{equation}
    - 10^{-4} \Delta R_i + R_i
    = S_i \quad \text{ on } \Omega_{\rm CSF}
    \qquad \text{ and } \quad
    \nabla R_i \cdot \bm n = 0, \quad \text{ on } \partial \Omega_{\rm CSF} .
\end{equation}
\end{enumerate}
Considering the uncertainty associated with the validity of
simplifying assumptions, the resulting estimates of the cardiac and
respiratory dispersion factors $R_c$ and $R_r$ 
should be viewed as heuristic rather than absolute.

%% \mar{To examine cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm LV}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model.}

%% % respiratory pulsatility
%% \mar{Employing the same methodology, we estimate the respiratory contribution to dispersive mixing by computing the CSF flow field driven by an inflow of $0.121\,$ml/s \cite{liu2024using} and $1\,$ml/s \cite{gutierrez2022effect} across the lateral ventricular wall and the arachnoid mater, respectively.}

%% Taking the large spatial differences of pulsatile CSF flow into account, we estimate a spatially varying diffusion enhancement factor $R$ in the CSF-filled space $\Omega_{\rm CSF}$ by combining a computational model of cardiac- and respiration-induced pulsatile CSF flow with theoretical estimates for shear-augmented (Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion}. These estimates are are derived under simplifying assumptions on the problem geometry, the flow field and concentration profiles \cite{watson1983diffusion, keith2019dispersion}. Therefore such an estimate is heuristic; however, our computational results qualitatively agree with observed data.

%% Similar to the case of CSF flow driven by production introduced in \Cref{sec:csf_flow}, we employ a Stokes flow model to predict CSF velocities corresponding to the peak cardiac and respiratory blood inflow rates. In particular, we enforce a uniformly distributed inflow of $u^{\rm in}_{\rm AM}$ and $u^{\rm in}_{\rm LV}$ across the arachnoid mater ($\Gamma_{\rm AM}$) and lateral ventricular surfaces ($\Gamma_{\rm LV}$) respectively, mimicking the effect of reduced CSF volume due to blood inflow. We set $u^{\rm in}_{\rm AM} = 6\,$ml/s \cite{causemann2022human,baledent2014imaging} and $u^{\rm in}_{\rm LV}=0.31\,$ml/s \cite{vinje2019respiratory} in the cardiac case, and $u^{\rm{in}}_{\rm{AM}} = 1\,$ml/s \cite{gutierrez2022effect} and $u^{\rm in}_{\rm{LV}}=0.121\,$ml/s \cite{liu2024using} in the respiratory case, respectively.
%% We prescribe a no-slip boundary condition on the pial surface $\Gamma_{\rm Pia}$ and zero-traction on the spinal boundary $\Gamma_{\rm SSAS}$, making the spinal compartment the only outflow route of CSF. Thus, we compute the pulsatile flow fields by augmenting the Stokes flow model (\eqref{eq:momentum_equation} and \ref{eq:divergence_equation}) with the following boundary conditions:
%% \begin{equation}\label{eq:cardiac_flow_bcs}
%%     \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
%%     \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
%%     \bm u = \frac{u^{\rm in}_{\rm LV} \cdot \bm n}{|\Gamma_{\rm LV}|} \text{ on } \Gamma_{\rm LV}; \quad
%%     \bm u = \frac{u^{\rm in}_{\rm AM} \cdot \bm n}{|\Gamma_{\rm skull}|} \text{ on } \Gamma_{\rm AM}.
%% \end{equation}

%% The obtained pressure fields $p_{\rm cardiac}$ and $p_{\rm resp}$ only account for viscous forces, albeit cardiac and respiratory induced CSF flow is known to be unsteady and substantially impacted by transient inertial forces. Assuming an oscillatory flow with either the heartbeat frequency $\omega_{\rm cardiac} = 2 \pi$ or respiratory frequency $\omega_{\rm resp} = 0.5 \pi$, a mean gap width of the CSF-filled spaces of $2h=3$\,mm and CSF density $\rho=10^3\,\text{kg/m}^3$, the squared Womersley number $\alpha_i^2 = \frac{h^2 \omega_{i} \rho}{\mu}$ becomes $\alpha_{\rm cardiac}^2 = 20.2$ and $\alpha_{\rm resp}^2=5.05$, which is unsteady. Based on theoretical considerations on the ratio of oscillatory flow to steady flow impedances in a tube, we account for inertial forces in the pulsatile flow by upscaling the computed viscous pressure field by a factor of $1 + \alpha^2 / 8$ (see \cite{van1998cardiovascular}, chapter. 4.3). Thus, the upscaled pressure fields become $p^{\rm up}_{i}= (1 + \alpha_{i}^2 / 8)p_i \text{ for } i = \{\mathrm{cardiac}, \mathrm{resp}\}$.

%% Further assuming unsteady dispersion, we follow Sharp et al \cite{keith2019dispersion} and estimate each enhancement factor $R_{\rm cardiac}$ and $R_{\rm resp}$ from the non-dimensionalized pressure gradient 
%% $$R_{i} \approx dP_{i}^2 / \alpha_{i}^3, \quad  \text{where} \quad  dP=\frac{|\nabla p^{\rm up}_{i}|}{\omega_{i} \mu / h} \quad \text{for} \quad i = \{\mathrm{cardiac}, \mathrm{resp}\}.$$
%% Finally, we account for the non-local nature of the dispersion mechanism and smooth the resulting field by solving a Helmholtz problem with $dP+i^2/\alpha_i^3$ as the right-hand side
%% \begin{equation}
%% - 10^{-4} \Delta R_i + R_i = dP_i^2/\alpha_i^3 \quad \text{ on } \Omega_{\rm CSF} \qquad \text{ and } \quad \nabla R_i \cdot \bm n = 0, \quad \text{ on } \partial \Omega_{\rm CSF}
%% \end{equation}
%% to obtain the locally averaged, spatially varying diffusion enhancement factor $R_i$. The diffusion enhancement factor $R$ in $D^{\mathrm{eff}} = D^{\mathrm{Gad}}( 1 +R)$ is then estimated by $R \approx R_{\rm{cardiac}} + R_{\rm{resp}}$.
